GMSE: an R package for generalised management strategy evaluation

### 2017-11-02 notebook/ms/ms.Rmd
Comments from @ninbrm and @ErlendNilsen included

GMSE builds off of the MSE framework (Figure 1), and the primary function `gmse` runs simulations using four default submodels. (1) A population of discrete resources with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty. This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management. (2) Observation is modelled in one of four ways: resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and resource counting across the whole landscape either one linear transect or one rectangular block at a time (during which resources might move). Sampling error from all observation types generates a range of uncertainties that depend on monitoring effort. (3) Managers analyse data collected from observations to estimate resource abundance, then compare this estimate with their pre-defined target abundance. Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for user actions on the resource (e.g., culling, scaring, etc.) that minimise deviation from the target abundance, as informed by the predicted consequences of each action on resource abundance and user action histories. After a suitable policy is found, (4) users perform actions that affect resources or landscape cells. Users respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints. Once each user has found an adaptive strategy, user actions affect resources and landscape cells, feeding back into the resource sub-model.

@@

Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders). Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step. Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017]. In finding strategies, GMSE models this trade-off by setting a fixed budget for managers and users. Allocations from a manager's budget can be used to increase the cost it takes a user to perform an action (i.e., 'policy'), and allocations from a user's budget can be used to perform the action at the cost set by the manager. Hence, stakeholders can have incomplete control over resource use and express competing management objectives.

In each new call of the genetic algorithm, a unique population of managers or users with random strategies is temporarily initialised. In each generation of the genetic algorithm, these strategies crossover and mutate; when this results in strategies that are over-budget, expenditures are iteratively decreased at random until budget constraints are satisfied. A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013]. The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fitness of the fittest strategy between the current and previous generation is below some threshold. The highest fitness strategy in the population then becomes the stakeholder's new strategy.

@@

|  res_consume    |  0.5    | Proportion of a landscape cell's value reduced by the presence of a resource in a time step      |

@@

|  minimum_cost   |  10     | The minimum cost of a stakeholder performing any action                                          |
|  user_budget    |  1000   | A stakeholder's budget per time step for performing any number of actions                        |

@@

|  tend_crops     |  FALSE  | Stakeholders can increase landscape cell values                                                  |

@@

|  kill_crops     |  FALSE  | Stakeholders can decrease landscape cell values to zero                                          |
|  stakeholders   |  4      | Number of stakeholders in the simulation                                                         |
|  land_ownership |  FALSE  | Stakeholders own land and increase utility indirectly from landscape instead of resource use     |

@@

|  public_land    |  0      | Proportion of land that is public (un-owned by stakeholders) if land_ownership = TRUE            |

@@

**Table 1:** Select parameter values for initialising generalised management strategy evaluation simulations

@@

Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of user land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops [e.g., @Madsen2017]. Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. We consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size. Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl for five farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

@@

The `gmse_apply` function simulates a single GMSE time step, and therefore must be looped for simulations over multiple time steps.

>>>>

GMSE builds off of the MSE framework (Figure 1). The workhorse function `gmse` runs simulations using four predefined submodels, which can be parameterised to fit various case studies; more tailored submodels can also be defined using the `gmse_apply` function (see below). (1) A population of discrete resources with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty. This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management. (2) Observation is modelled in one of four ways: resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and resource counting across the whole landscape either one linear transect or one rectangular block at a time (during which resources might move). Sampling error from all observation types generates a range of uncertainties that depend on monitoring effort. (3) Managers analyse data collected from observations to estimate resource abundance, then compare this estimate with their pre-defined target abundance. Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for user actions on the resource (e.g., culling, scaring, etc.) that minimise deviation from the target abundance, as informed by the predicted consequences of each action on resource abundance and user action histories. After a suitable policy is found, (4) users perform actions that affect resources or landscape cells. Users respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints. Once each user has found an adaptive strategy, user actions affect resources and landscape cells, feeding back into the resource sub-model.

@@

Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders). Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step (see Supporting Information 1 for details). Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017]. In finding strategies, GMSE models this trade-off by setting a fixed budget for managers and users. Allocations from a manager's budget can be used to increase the cost it takes a user to perform an action (i.e., 'policy'), and allocations from a user's budget can be used to perform the action at the cost set by the manager. Hence, stakeholders can have incomplete control over resource use and express competing management objectives.

In each new call of the genetic algorithm, a unique population of managers or users with random strategies is temporarily initialised. In each generation of the genetic algorithm, these strategies crossover and mutate; when this results in strategies that are over-budget, expenditures are iteratively decreased at random until budget constraints are satisfied. A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013]. The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fitness of the fittest strategy between the current and previous generation is below some threshold. The highest fitness strategy in the population then becomes the stakeholder's new strategy. Overall, a single call of the genetic algorithm thereby simulates the process of thinking and decision-making for one manager or user.

@@

|  res_consume    |  0.5    | Pr. of a landscape cell's value reduced by the presence of a resource in a time step             |

@@

|  minimum_cost   |  10     | The minimum cost of a user performing any action                                                 |
|  user_budget    |  1000   | A user's budget per time step for performing any number of actions                               |

@@

|  tend_crops     |  FALSE  | Users can increase landscape cell values                                                         |

@@

|  kill_crops     |  FALSE  | Users can decrease landscape cell values to zero                                                 |
|  stakeholders   |  4      | Number of users in the simulation                                                                |
|  land_ownership |  FALSE  | Users own land and increase utility indirectly from landscape instead of resource use            |

@@

|  public_land    |  0      | Proportion of land that is public (un-owned by users) if land_ownership = TRUE                   |

@@

Table 1: Select parameter values for initialising generalised management strategy evaluation simulations

@@

Here we illustrate the usefulness of GMSE by considering the case study of a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers attempt to keep the abundance of waterfowl at a target level, while farmers attempt to minimise the damage inflicted on their crops [e.g., @Madsen2017]. Using GMSE, we can simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. We consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size. Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl for five farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

@@

The `gmse_apply` function simulates a single GMSE time step, and therefore must be looped for simulations over multiple time steps. Within loops, GMSE arguments can be redefined to simulate changing conditions (e.g., change in policy availability or stakeholder budgets, see Supporting Information 2).
